---
title: "Spider Diagrams"
author: "Jade Bowers"
date: "1/13/2021"
output:
  pdf_document: default
  html_document: default
---

## **Abstract**

```{r setup, include=FALSE}
#this code chunk only pertains to the function of the Markdown document and is not necessary for running the code itself in an R script

knitr::opts_chunk$set(echo = TRUE)
```

## **Introduction**

The R Markdown document is designed as a template for open source use and contains five code chunks necessary for recreate a spider diagram on your own in a separate R script.

## **The Code**

### *1. Required package(s)*

```{r packages, echo=TRUE, results='hide', message=FALSE, warning=FALSE}
library(tidyverse)
library(tidyr)
library(dplyr)
library(ggpubr)
library(ggthemes)
```

Sequencing the elements creates an index that plots the elements in the correct order on the x axis rather than R's default alphabetical order.
The character vector, *traditional*, is a vector of the traditional trace elements plotted on a multi-element diagram in the common order.
This vector, or a new vector, can be modified to include or remove any elements based on your plotting needs.
Just delete them from the character vector.
Their sequencing will update based on the changes when you run line 8 of the code.

```{r creating sequencing of elements}
#list of traditional TEs for filtering
traditional = c("Rb","Ba", "Th" ,"U","Nb", "La", 
                "Ce","Pb","Sr", "Nd","Zr","Hf", 
                "Sm", "Eu", "Gd", "Tb", "Dy", "Y", 
                "Ho", "Er", "Yb", "Lu")  
#creates a df to join with your data
TE_seq = data.frame(elements = traditional,
                    seq = seq(1:length(traditional)))
```

```{r load data, results='hide', message=FALSE}
example_data <- read_csv("../data/spider_example_data.csv")
example_data <- example_data %>% 
  select(sample, Cs:Lu) %>% 
  gather(key = "element", value = "value", -sample)
sequenced_data <- example_data %>% inner_join(TE_seq, by = c("element" = "elements"))
```

Multi-element diagrams exist on a log-scale, next we create the base ggplot object that you will plot your data on top of.

```{r}
spider_diagram <- ggplot() +
 aes(x = seq, y = value) +
  scale_x_continuous(name = "", 
                     breaks = sequenced_data$seq, 
                     labels =sequenced_data$element) +
  ylab("Sample") +
  scale_y_log10(breaks = c(0.01, 0.1, 1, 10, 100),
                labels = c(0.01, 0.1, 1, 10, 100)) +
  theme_stata()
```

Add your data to the base plot.
Make sure your data has a "sample" column to group your data buy.
Without it, R will connect your data in its default hierarchical order and the plot will not look correct.

```{r}
spider_diagram +
  geom_line(data = sequenced_data, aes(group = sample, color = sample)) +
  guides(col = guide_legend(nrow = 3)) +
  theme(axis.text.y = element_text(angle = 0))
```
