---
title: "Error Propagation for Trace Element Ratios"
author: "Jade Bowers"
date: "2023-07-25"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Error Propagation for Trace Element Ratios

Often, uncertainty is only plotted on element-element bivariate diagrams.
Including error bars on bivariates that have an element-element ratio on one of the axes requires error propagation through the ratio calculation.

Example: You want to include error bars on your plot of La/Sm

```{r}
errorMat <- matrix(nrow = 1, ncol = 3)
errorMat[,1] <- 0.0126
errorMat[,2] <- 0.0124
rownames(errorMat) <- "oneSig"
colnames(errorMat) <- c("La", "Sm", "LaSm")

data <- matrix(nrow = 1, ncol = 2)
data[,1] <- 12
data[,2] <- 2.63
rownames(data) <- "sample"
colnames(data) <- c("La", "Sm")
```

Simple function for propagating error through calculation of an element-element ratio.

```{r}
error_prop <- function(x, y, delta_x, delta_y) {
  # Inputs:
    ##  x = concentration of numerator element in your sample
    ##  y = concentration of denominator element in your sample
    ##  delta_x = one sigma uncertainty of numerator element
    ##  delta_y = one sigma uncertainty of denominator element
  # Outputs:
    ##  delta_z = one sigma uncertainty of trace element ratio
  
    delta_z = (x/y) * sqrt(((delta_x/x)^2) + ((delta_y/y)^2))

  return(delta_z)
}
```

```{r}
errorMat["oneSig", "LaSm"] <- error_prop(x = data[1,"La"], 
                                       y = data[1,"Sm"], 
                                       delta_x = errorMat["oneSig", "La"], 
                                       delta_y = errorMat["oneSig", "Sm"])
print(errorMat)
```
